---
title: "class_10_lab"
author: "Tessa Sterns PID: A18482353"
format: pdf
toc: true
---

## Importing candy data

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)

head(candy)
```

```{r}
flextable::flextable(head(candy))
```

>Q1. How many different candy types are in this dataset?

There are `r nrow(candy)` types of candy in this dataset.

>Q2. How many fruity candy types are in the dataset?

There are `r sum(candy$fruity)` types of fruity candy in this dataset.

## Overall Candy Rankings

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in this dataset is Milkyway Midnight
with a win percent of `r candy["Milky Way Midnight", ]$winpercent`.

>Q4. What is the winpercent value for “Kit Kat”?

Kit Kat has a winpercent of `r candy["Kit Kat", ]$winpercent`.

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

Tootsie Roll Snack Bars has a win percent of `r candy["Tootsie Roll Snack Bars", ]$winpercent`.

```{r}

skimr::skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes the win percent column is in a whole number percent(ie 65.4%) whereas the other columns like percent sugar are in a decimal point percent (ie 0.923).

>Q7. What do you think a zero and one represent for the candy$chocolate column?

They are the binary representation of T/F 1 would be true/ has chocolate, and 0 would be false/no chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
aes(x = winpercent) +
  geom_histogram(binwidth = 6) +
  theme_minimal()
```

>Q9. Is the distribution of winpercent values symmetrical?

No it it slightly right skewed with the normal distribution being centered closer to 40%.

>Q10. Is the center of the distribution above or below 50%?

Below 50% it is around 40%.
`r summary(candy$winpercent)`

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
library(tidyverse)
j.choc <- candy %>% 
  filter(chocolate == 1) 
mean(j.choc$winpercent)

j.fruit <- candy %>%
  filter(fruity == 1)
mean(j.fruit$winpercent)
```
Chocolate has a higher average win percent `r mean(j.choc$winpercent)` than fruity candy `r mean(j.fruit$winpercent)`.

>Q12. Is this difference statistically significant?

```{r}
t.test(x = j.choc$winpercent, 
  y = j.fruit$winpercent)
```
With a p-value of 2.871e-08 we accept the null hypothesis of the two-sample ttest. This indicates that there is a statistical difference between chocolate and fruity candy with chocolate being more popular.

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

The five least liked candies are; `r candy %>%
  arrange(winpercent) %>% 
  head(5) %>%
  row.names()`.

> Q14. What are the top 5 all time favorite candy types out of this set?

The top five most liked candies are;  `r candy %>%
  arrange(winpercent) %>% 
  tail(5) %>%
  row.names()`.
  
>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(x= winpercent, rownames(candy)) +
  geom_col()
```
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(x= winpercent, y= reorder(rownames(candy), winpercent)) +
  geom_col() +
  theme(axis.text.y = element_text(size = 3))
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "lightpink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + 
   theme(axis.text.y = element_text(size = 3))
```
>Q17. What is the worst ranked chocolate candy?

Sixlets

>Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=2, max.overlaps = 10) +
  theme_minimal()
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The most expensive candies are `r rownames(head( candy[ord,c(11,12)], n=5 ))` and the least liked is `r rownames(head( candy[ord,c(11,12)], n=1) )`.

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy),     pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
  xend = 0), col=my_cols) +
    geom_point() +
  theme(axis.text.y = element_text(size = 3))
```

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

>Q23. Similarly, what two variables are most positively correlated?

chocolate and bar or chocolate and win percentage

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
library(plotly)
## note plotly can make interactive plots but wont work for a pdf

p <- ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

Examining variable "loadings" or contributions of original variables to new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. This makes sense the way many fruity candies are packaged like skittles or starburst you get a lot of little candies. Additionally things like lolipops and suckers are more often fruit flavored.
