---
title: "Class_17_Hw"
author: "Tessa Sterns PID:A18482353"
format: pdf
---

```{r}
my_data <- read.table("asthma_results.txt")

library(ggplot2)

summary(my_data)
```

```{r}
library(dplyr)

AA <- my_data %>% filter(geno == "A/A")

AG <- my_data %>% filter(geno == "A/G")

GG <- my_data %>% filter(geno == "G/G")
```

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The A/A genotype has `r nrow(AA)` samples with a median expression of `r median(AA$exp)`. The A/G genotype has `r nrow(AG)` samples with a median expression of `r median(AG$exp)`.The G/G genotype has `r nrow(GG)` samples with a median expression of `r median(GG$exp)`.

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
my_plot <- ggplot(my_data) + 
  aes(x = geno, y = exp,
      fill = geno) +
  geom_boxplot() +
  labs(x = "genotype", y = "expression") +
  theme_minimal() +
  theme(legend.position = "FALSE")
my_plot
```
The SNP does effect expression levels with A/A showing the highest expression. Between AA and GG there is about a 12 point decrease in expression. These results indicate that the SNP allele of ORMDL3 can cause a change in expression levels of the protein product wich can lead to and increased risk of childhood asthma.  
