---
title: "R Functions Lab Class_06"
author: "Tessa Sterns PID:18482353"
format: pdf
---

Every function in R contain at least 3 things:

 - A **name** that we choose.
 - Input **arguments**, multiple possible seperated by a comma.
 - The **body**, lines of R code that do the work of the function.

My first lil function;

```{r}
add <- function(x, y = 5) {x + y}

add(10)
add(10,20)
add(c(1,2,3), 2)
```

## The Second Function!
A sequence generator 
The `sample()` function can be useful in this
```{r}
sample(1:10, size = 3)
sample(c("A", "G", "T", "C"), size = 20, replace = TRUE)
DNA_Generator <- function( b=10 ) {
  v <- sample(c("A", "G", "T", "C"), size = b, replace = TRUE) 
  cat("Good Job\n")
  return(v)}
DNA_Generator(30)
DNA_Generator()
```
```{r}
s <- DNA_Generator(20)
s
```

I want the option to return a single element vector "TATTAA"
The `paste0()` with collapse = "" will reduce a vector down to a single element. 
```{r}
DNA_Generator2 <- function( b=10, type = c("multi", "fasta")) {
  type <- match.arg(type)
  v <- sample(c("A", "G", "T", "C"), size = b, replace = TRUE) 

  ## single element vector
  cat("Good Job\n")
  if (type == "fasta") {
    return(paste0(v, collapse = ""))
  } else {
    # Return the vector of individual bases
    return(v)
  }
}
DNA_Generator2(40, "fasta")

```

Protein sequence generator

```{r}
Protein_Gen <- function( b=10, type = c("PolyP", "multi")) 
  ## make the defult return the first argument in type()
  {
  type <- match.arg(type)
  
  AA <- sample(c("A","R", "N", "D", "Q", "H", "I",
                 "E", "G", "T", "C", "L", "K", "M",
                 "F", "P", "S", "W", "Y", "V"),
               size = b, replace = TRUE) 
 
  ## single element vector
 
  if (type == "multi") {
    return(AA)
  P} else {
    # Return the vector of individual bases
    return(paste0(AA, collapse = ""))
  }
}
Protein_Gen(37, "multi")
Protein_Gen(65)
```

>Q. Generate random protein sequences between length 5 and 12 amino-acids.

```{r}
sapply(5:12, Protein_Gen)  |> cat(sep = "\n")
```

> **Key-Point**: writting functions in R is doable but not always intuitive. Starting with a working portion of code and using LLM tools to modify it can generate a more robust working function.
